---
title: "Poisson Panel Data Model with STAN"
format: 
  html:
    embed-resources: true
---

Suppose that $Y_{it} \sim \text{Poisson}(\mu_{it})$ where 
$$
\begin{align*}
\log(\mu_{it}) &= \alpha_i + \beta X_{it} + \gamma Z_i \\
\alpha_i &= \bar{\alpha} + \tau U_i \\
U_i &\sim \text{Normal}(0, 1)
\end{align*}
$$
and $X_{it}$ is potentially correlated with $U_i$. First we'll simulate some data:
```{r}
Nt <- 3
Ni <- 250
N <- Nt * Ni
alpha <- (-2)
beta <- 1
gamma <- (-0.5)

set.seed(1693)

# Vector of individual ids in the full panel 
id <- rep(1:Ni, each = Nt)

# Variables that are fixed over time 
U <- rnorm(Ni, 1.5) # unobserved
Z <- rnorm(Ni) # observed

# Variables that vary across i and t
X <- rnorm(N, mean = U[id])
mu <- exp(alpha + beta * X + gamma * Z[id] + U[id])
Y <- rpois(N, mu)
```

Next we'll compile a STAN program that uses the "fast-and-dirty" Mundlak approach:
$$
\begin{align*}
Y_{it} &\sim \text{Poisson}(\mu_{it})\\
\log(\mu_{it}) &= \alpha_i + \beta X_{it} + \gamma Z_i + \delta \bar{X}_i\\
\alpha_i &= \bar{\alpha} + \tau U_i\\
U_i &\sim \text{Normal}(0, 1)\\
\bar{\alpha}, \beta, \gamma, \delta &\sim \text{Normal}(0, 1)\\
\tau &\sim \text{Exponential}(1)
\end{align*}
$$
Notice that we **require a sign normalization** for $\tau$ since $U_i$ is a latent variable.
Next we'll compile the STAN program:
```{r}
#| warning: false
#| message: false
library(cmdstanr)
poisson_panel_model  <- cmdstan_model('poisson-panel-simple.stan')
```
And then we'll run it on the simulation data:
```{r}
#| warning: false
#| message: false
library(tidyverse)
dat <- list(X = X, Y = Y, Z = Z, id = id, N = N, Ni = Ni)

dat$Xbar <- tibble(X = X, g = id) |> 
  group_by(g) |> 
  summarize(Xbar = mean(X)) |> 
  pull(Xbar)

poisson_panel_fit <- poisson_panel_model$sample(
  data = dat,
  seed = 123,
  chains = 4,
  parallel_chains = 4,
  refresh = 500
)

poisson_panel_fit$summary(variables = c('abar', 'beta', 'gamma', 'delta', 'tau')) |>
                            knitr::kable(digits = 2)
```
The results look good! Now let's try the "full luxury Mundlak" approach.




We will model this as follows:
$$
X_{it} \sim N(\eta_i, \nu), \quad \eta_i \equiv \lambda + \kappa U_i
$$
We either need to normalize $\kappa$ to be positive or normalize $\tau$ to be positive. Our previous experiments with a linear panel model suggested that it was a better choice to normalize $\kappa$: this resulted in a much more efficient sampler. We're not sure why, but we'll use the same approach here. It's presumably something to do with the geometry of the problem. 
